Fluent statistical computing interfaces for biological data analysis

### 2020-08-03 Rmd/05-embeddings.Rmd
addressing typos in #9

A complementary approach for visualizing structure in high dimensional data is 
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as a
animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
nature of the tour, user interaction is  important for controlling and 
exploring the visualisation: the tour has been used previously by 
@Wickham2015-cx and exploring statistical model fits and by @Buja1996-fk for 
exploring the space of factorial experimental designs.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

@@

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

Like, when using other DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections.  In practice, we first sphere 
our data via principal components to reduce dimensionality of $X$ prior to 
running the tour. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence of 'interesting' 
projections as quantified by an index function [@Cook1995-bi]. While our software, 
**liminal** is able to  visualise guided tours, our focus in the case studies 
uses the grand tour to see global structure in the data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatter plot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flip book" 
animation. Like with a flip book, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in **liminal**
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

@@

The next section steps through case studies cases of our approach using simulated
and real data.

@@

Next explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.

>>>>

A complementary approach for visualizing structure in high dimensional data is 
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as an
animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic 
nature of the tour, user interaction is  important for controlling and 
exploring the visualisation: the tour has been used previously by 
@Wickham2015-cx for exploring statistical model fits and by @Buja1996-fk for 
exploring the space of factorial experimental designs.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR method that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

@@

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a t-distribution with one degree of freedom (Cauchy kernel):

@@

Like, when, using other DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections.  In practice, we first sphere 
our data via principal components to reduce dimensionality of $X$ prior to 
running the tour. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence of 'interesting' 
projections as quantified by an index function [@Cook1995-bi]. While our software, 
**liminal** is able to  visualise guided tours, our focus in the case studies 
uses the grand tour to see global structure in the data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near, or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ and the tour is visualised as
an animated scatter plot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flip book" 
animation. Like with a flip book, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in **liminal**
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

@@

The next section steps through case studies of our approach using simulations
and an application to single cell RNA-seq data.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.

### 2020-08-03 Rmd/06-conclusion.Rmd
addressing typos in #9

The **superintronic** software described in chapter \@ref(ch:intron) has
been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
data. Here, we again show the strengths of providing a long-form representations
of genomics data (in this case coverage vectors). By leveraging **plyranges**
we were then able to create a set of data descriptors that we could link back
to the raw data to discover genes thought to be associated with a real biological 
signal. An interesting extension to this work would be applying it to single
cell and long-read based transcriptomics data, where scalability and much
larger design matrices would be come an issue. The **superintronic** package
is available to download from https://github.com/sa-lee/superintronic.

@@

A limitation of the grammar as we have implemented it in 
**plyranges** is lack of scalability and computational speed for data sets
that do not fit in memory. We attempted several techniques for performing 
delayed operations over range-based data, however a more general approach
that allows for data stored on the cloud or in scientific data formats
like HDF5 that leverage existing Bioconductor frameworks would be useful.
We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an
analyst is able to do some very complex data transformations and re-sampling 
procedures via casting results into _GRanges_ object. However, it is unclear 
whether the semantics of our grammar can be extended to data that can not be 
easily reshaped into long form tidy representations. Moreover, further work
is required to explore the design space of grammars for data 
transformations and grammars for graphics when the data large, multifaceted
and non-rectangular.

>>>>

The **superintronic** software described in chapter \@ref(ch:intron) has
been used in @Lee2019-mf to disentangle and view intron signal in RNA-seq
data. Here, we again show the strengths of providing a long-form representations
of genomics data (in this case coverage vectors). By leveraging **plyranges**
we were then able to create a set of data descriptors that we could link back
to the raw data to discover genes thought to be associated with a real biological 
signal. An interesting extension to this work would be applying it to single
cell and long-read based transcriptomics data, where scalability and much
larger design matrices would become an issue. The **superintronic** package
is available to download from https://github.com/sa-lee/superintronic.

@@

A limitation of the grammar as we have implemented it in 
**plyranges** is lack of scalability and computational speed for data sets
that do not fit in memory. We attempted several techniques for performing 
delayed operations over range-based data, however a more general approach
that allows for data stored on the cloud or in scientific data formats
like HDF5 that leverage existing Bioconductor frameworks would be useful.
We showed in chapters \@ref(ch:fluentGenomics) ad \@ref(ch:intron) that an
analyst is able to do some very complex data transformations and re-sampling 
procedures via casting results into _GRanges_ object. However, it is unclear 
whether the semantics of our grammar can be extended to data that can not be 
effeciently reshaped into long form tidy representations. Moreover, further work
is required to explore the design space of grammars for data 
transformations and grammars for graphics when the data are large, multifaceted
and non-rectangular.
